Safe Transferable Regions

### 2017-01-09 paper/misc.tex
type inference minor updates

\section{Other Aspects}

\paragraph{Modularity Aspects of Type Inference.}
The type inference algorithm, as presented, traverses the entire program to
generate the set of constraints, which are solved en masse, using an iterative
fixed point computation. However, the type inference can be realized in a
modular and compositional fashion, subject only to the restrictions imposed
by recursion.

>>>>

% \subsection{Other Aspects}

\subsection{Modularity Aspects of Type Inference}

The type inference algorithm, as presented, traverses the entire program to
generate the set of constraints, which are solved en masse, using an iterative
fixed point computation. However, the type inference can be realized in a
modular and compositional fashion, subject only to the restrictions imposed
by recursion.

### 2017-01-09 paper/type-inference.tex
type inference minor updates

We later on establish the soundness of the type inference algorithm,
and the completeness of constraint-solving step.

@@

\begin{tabular}{ccc}
\begin{minipage}{0.33\linewidth}
\begin{codejava}
class Pair $\extends$ $\ObjZ$ {
  $\ObjZ$ fst;
  $\ObjZ$ snd;
}
\end{codejava}
\end{minipage}
&
$\Rightarrow \; \;$
&
\begin{minipage}{0.6\linewidth}
\begin{codejava}
class Pair $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$ {
  $\ObjZ \langle \rho_1 \rangle$ fst;
  $\ObjZ \langle \rho_2 \rangle$ snd;
}
\end{codejava}
\end{minipage}
\end{tabular}

@@

\begin{tabular}{ccc}
\begin{minipage}{0.33\linewidth}
\begin{codejava}
class List $\extends$ $\ObjZ$ {
  $\ObjZ$ data;
  List next;
}
\end{codejava}
\end{minipage}
&
$\Rightarrow \; \;$
&
\begin{minipage}{0.65\linewidth}
\begin{codejava}
class List $\langle \rho_0, \rho_1 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$ {
  $\ObjZ \langle \rho_1 \rangle$ data;
  $\C{List} \langle \rho_0, \rho_1 \rangle$ next;
}
\end{codejava}
\end{minipage}
\end{tabular}

@@

\emph{Type-Parametric Classes}.
% Type parameters of classes are handled as follows.
% Consider a type-parametric class \C{C $\langle$ T $\extends$ B $\rangle$}.
The type parameter \C{T} of a class \C{C} is instantiated as $\C{T}@\rho$ using a single region parameter $\rho$. 
(This can be extended to use the bound specified for \C{T}, if any. We do not
discuss this due to space restrictions.)
% number of region parameters
% that its bound \C{B} has. If no bound is specified for \C{T}, the bound is taken
% to be \C{Object}, and \C{T} is instantiated with one region parameter.

@@

Recall that a \emph{region variable} is introduced to represent an unknown actual region parameter
of a method invocation or object allocation, which will be bound to a region constant by the end of the
type inference.
A region constant may be 
(a) a \emph{formal region parameter} of a class or method, or
(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct.

A predicate variable $\varphi$ is introduced to represent, \eg, the unknown precondition of a method.
The constraint-solver will end up binding it to a \emph{region-constraint} $\phi$
over a set of fixed formal region parameters.
%
%% We will use $\rho_1 = \rho_2$ as shorthand for $\rho_1 \outlives \rho_2 \conj \rho_2 \outlives \rho_1$.
%% We will also use a set $S$ of primitive constraints of the form $\rho_1 \outlives \rho_2$
%% as shorthand for $\conj_{\phi \in S} \phi$.
%% 
% We will use the term \emph{validity constraint} to denote an entailment constraint
% of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.
%
Our constraints also make uses of \emph{pending substitutions} $F$:
A pending substitution serves to bind formal region parameters in $\varphi$ to the actual region parameters
used in a particular context:
E.g., in the validity constraint $\isvalid{\rgn_1 \outlives
\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, the pending substitution
is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$.
% Any concrete formula (call it
% $\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
% $\varphi$ if and only if the formula obtained by substituting $\rgn_1$
% and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
% deducible from $\rgn_1 \outlives \rgn_2$.

@@

We use \emph{well-formedness constraints} of the form $\rho \in \rhoenv$
to restrict the domain of unification for a region variable ($\rho$)
to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of regions in scope,
and  well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$ to
restrict the domain of a predicate variable ($\varphi$) to the set of all possible
constraint formulas over a fixed set of  regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope.

@@

Fig.~\ref{fig:constraint-gen-0} illustrates this for selected language constructs.
(The appendix contains the remaining constraint generation rules.)
These rules use the same context as the corresponding typing judgment in
Fig.~\ref{fig:fb-staticsem}, except that this is generalized to permit the
use of region variables and predicate variables.
Symbols $\A$ and $\env$ retain their meaning, modulo this extension.
Thus, the component $\phicx$ of $\A$ will now be a symbolic expression
of the form $\varphi \conj \phi$, where $\varphi$ is the predicate variable
representing the undetermined precondition of the method being analyzed.
The algorithm proceeds top-down, analyzes an expression $e$ in a context
$\stdcontext$, and returns a type $\tau$ and a set of constraints $C$
(expressed in the rules as $\exprok {\stdcontext} {e} {\tau} {C}$ ),
indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.

Note that some of the conditions in the antecedents are checked at constraint-generation
time, while some are encoded as constraints (to be resolved by the constraint-solver).
E.g., we assume that the region constants introduced by
\C{open}
and
\C{letregion}
constructs are unique (and the elaboration phase ensures this by alpha-renaming
them as needed). Thus, the precondition $\rho \not\in \Delta$ in rules \textsc{LetRegion}
and \textsc{Open} merely validates this at constraint-generation time.

@@

Let $\consOf{q}$ denote the set of constraints generated from an elaborated program $q$.
The following theorem states that constraint-generation is sound:
\begin{theorem}
\label{thm:constraint-generation-sc}
Let $C = \consOf{q}$.
An assignment $\soln$ (for the region and predicate variables in $q$)
satisfies $C$ only if $q[\soln]$ is well-typed.
\end{theorem}

@@

\begin{tabular}{ccc}
\begin{minipage}{0.32\linewidth}
\begin{codejava}
class List $\extends$ $\ObjZ$
{
  $\ObjZ$ data;
  List next;
}
\end{codejava}
\end{minipage}
&
$\Rightarrow  \; \; \;$
&
\begin{minipage}{0.65\linewidth}
\begin{codejava}
class List $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$
{
  $\ObjZ \langle \rho_1 \rangle$ data;
  $\C{List} \langle \rho_0, \rho_2, \rho_1 \rangle$ next;
}
\end{codejava}
\end{minipage}
\end{tabular}

>>>>

We later on establish the soundness of the type inference algorithm,
and the completeness of the constraint-generation and constraint-solving steps
(steps 2 and 3 above).

@@

\begin{tabular}{ccc}
\begin{minipage}{0.33\linewidth}
\begin{codejava}
class Pair $\extends$ $\ObjZ$ {
  $\ObjZ$ fst;
  $\ObjZ$ snd;
}
\end{codejava}
\end{minipage}
&
$\Rightarrow \; \;$
&
\begin{minipage}{0.5\linewidth}
\begin{codejava}
class Pair $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$ {
  $\ObjZ \langle \rho_1 \rangle$ fst;
  $\ObjZ \langle \rho_2 \rangle$ snd;
}
\end{codejava}
\end{minipage}
\end{tabular}

@@

\begin{tabular}{ccc}
\begin{minipage}{0.33\linewidth}
\begin{codejava}
class List $\extends$ $\ObjZ$ {
  $\ObjZ$ data;
  List next;
}
\end{codejava}
\end{minipage}
&
$\Rightarrow \; \;$
&
\begin{minipage}{0.5\linewidth}
\begin{codejava}
class List $\langle \rho_0, \rho_1 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$ {
  $\ObjZ \langle \rho_1 \rangle$ data;
  $\C{List} \langle \rho_0, \rho_1 \rangle$ next;
}
\end{codejava}
\end{minipage}
\end{tabular}

@@

\emph{Type-Parametric Classes}.
% Type parameters of classes are handled as follows.
% Consider a type-parametric class \C{C $\langle$ T $\extends$ B $\rangle$}.
The type parameter \C{T} of a class \C{C} is instantiated as $\C{T}@\rho$ using a single region parameter $\rho$. 
(This can be extended to use the bound specified for \C{T}, if any.)
% number of region parameters
% that its bound \C{B} has. If no bound is specified for \C{T}, the bound is taken
% to be \C{Object}, and \C{T} is instantiated with one region parameter.

@@

Recall that a \emph{region constant} may be either
(a) a \emph{formal region parameter} of a class or method, or
(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct.
A \emph{region variable} is introduced to represent an unknown \emph{actual} region parameter
of a method invocation or object allocation, and the constraint-solver, if successful, 
will bind each region variable to a region constant.
% by the end of the type inference.

A \emph{region-constraint} $\phi$ consists of a conjunction of \emph{outlives-constraints}
of the form $\rho_1 \outlives \rho_2$.
A predicate variable $\varphi$ is introduced to represent, \eg, the unknown precondition of a method.
The constraint-solver will end up binding it to a \emph{region-constraint} $\phi$
over a set of fixed formal region parameters.
%
%% We will use $\rho_1 = \rho_2$ as shorthand for $\rho_1 \outlives \rho_2 \conj \rho_2 \outlives \rho_1$.
%% We will also use a set $S$ of primitive constraints of the form $\rho_1 \outlives \rho_2$
%% as shorthand for $\conj_{\phi \in S} \phi$.
%% 
% We will use the term \emph{validity constraint} to denote an entailment constraint
% of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.
%
Our constraints also make uses of \emph{pending substitutions} $F$:
A pending substitution serves to bind formal region parameters in $\varphi$ to the actual region parameters
used in a particular context:
E.g., in the validity constraint $\isvalid{\rgn_1 \outlives
\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, the pending substitution
is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$.
% Any concrete formula (call it
% $\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
% $\varphi$ if and only if the formula obtained by substituting $\rgn_1$
% and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
% deducible from $\rgn_1 \outlives \rgn_2$.

@@

We use \emph{well-formedness constraints} of the form $\rho \in \rhoenv$
to restrict the domain of unification for a region variable ($\rho$)
to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of regions in scope,
and  well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$ to
restrict the domain of a predicate variable ($\varphi$) to the set of all possible
region-constraint formulas over a fixed set of  regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope.

@@

Fig.~\ref{fig:constraint-gen-0} illustrates this for selected language constructs.
(The appendix contains the remaining constraint generation rules.)
These rules use the same context as the corresponding typing judgment in
Fig.~\ref{fig:fb-staticsem}, except that this is generalized to permit the
use of region variables and predicate variables.
Symbols $\A$ and $\env$ retain their meaning, modulo this extension.
Thus, the component $\phicx$ of $\A$ will now be a \emph{symbolic expression}
of the form $\varphi \conj \phi$, where $\varphi$ is the predicate variable
representing the undetermined precondition of the method being analyzed.
The algorithm proceeds top-down, analyzes an expression $e$ in a context
$\stdcontext$, and returns a type $\tau$ and a set of constraints $C$
(expressed in the rules as $\exprok {\stdcontext} {e} {\tau} {C}$ ),
indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.

When we adapt a type-checking judgement rule to produce a corresponding
constraint-generation rule, we treat the antecedent conditions of the type-checking
rule in one of two ways. Many of these antecedent conditions are converted into
constraints which are accumulated in the set $C$. However, some of the antecedent
conditions are expected to be trivially satisfied and are expressed as
antecedent conditions in the constraint-generation rule as well.
% Note that some of the conditions in the antecedents are checked at constraint-generation
% time, while some are encoded as constraints (to be resolved by the constraint-solver).
E.g., our frontend ensures that the region constants introduced by \C{open}
and \C{letregion} constructs are unique (and the elaboration phase ensures this by alpha-renaming
them as needed). Thus, the precondition $\rho \not\in \Delta$ in rules \textsc{LetRegion}
and \textsc{Open} is expected to hold true at constraint-generation time
(and does not produce any constraints).

@@

Let $\consOf{q}$ denote the set of constraints generated from an elaborated program $q$.
The following theorem states that constraint-generation is sound and complete:
\begin{theorem}
\label{thm:constraint-generation-sc}
Let $C = \consOf{q}$.
An assignment $\soln$ (for the region and predicate variables in $q$)
satisfies $C$ iff $q[\soln]$ is well-typed.
\end{theorem}

@@

\begin{tabular}{ccc}
\begin{minipage}{0.32\linewidth}
\begin{codejava}
class List $\extends$ $\ObjZ$
{
  $\ObjZ$ data;
  List next;
}
\end{codejava}
\end{minipage}
&
$\Rightarrow  \; \; \;$
&
\begin{minipage}{0.5\linewidth}
\begin{codejava}
class List $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$
{
  $\ObjZ \langle \rho_1 \rangle$ data;
  $\C{List} \langle \rho_0, \rho_2, \rho_1 \rangle$ next;
}
\end{codejava}
\end{minipage}
\end{tabular}
